---
title: "From Spreadsheets to Systems"
subtitle: "A companion resource to the worksheet provided at ELC 2026"
author:
  - Madeline Ratoza, PT, DPT, PhD
  - Sam Pak, PT, DPT
  - Peyton D. Sykes, PT, DPT, EdD
  - Morgan Ascher, PT, DPT
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 2
    toc-title: "Explore this resource"
    embed-resources: true
    code-fold: true
    code-summary: "View the R code"
    code-copy: true
    title-block-banner: "#163b50"
    title-block-banner-color: "#ffffff"
    include-in-header:
      text: |
        <style>
        #title-block-header .quarto-title-banner {
          background: #163b50 !important;
          color: #ffffff !important;
        }
        #title-block-header .quarto-title-banner .title,
        #title-block-header .quarto-title-banner .subtitle {
          color: #ffffff !important;
          opacity: 1 !important;
        }
        body {
          color: #213944;
          line-height: 1.7;
        }
        main h1, main h2, main h3 {
          color: #163b50;
        }
        main h1 {
          margin-top: 2.5rem;
        }
        a {
          color: #146b74;
        }
        table {
          width: 100%;
          margin: 1.4rem 0;
        }
        thead th {
          background: #163b50 !important;
          color: #ffffff !important;
        }
        td, th {
          padding: 0.7rem 0.9rem !important;
        }
        tbody tr:nth-child(even) {
          background: #f2f7f8;
        }
        details {
          margin: 1.5rem 0;
          padding: 1rem 1.2rem;
          border: 1px solid #cbdde1;
          border-radius: 8px;
          background: #f7fafb;
        }
        summary {
          color: #163b50;
          font-weight: 600;
          cursor: pointer;
        }
        @media print {
          #quarto-margin-sidebar, .code-copy-button {
            display: none;
          }
          h1, h2, h3 {
            break-after: avoid;
          }
        }
        </style>
engine: knitr
execute:
  echo: true
  warning: false
  message: false
---

::: {.callout-note}
## Welcome back to your worksheet

You identified a recurring educational-data problem and a small change you could test. This companion resource helps you turn those ideas into a pilot plan.

Start with your worksheet. Use the sections below to clarify your approach, involve the right people, and decide what evidence would justify continuing.
:::

# Turn your worksheet into a pilot

A useful first test is small enough to complete and specific enough to learn from. Choose one course, cohort, report, or review cycle.

## 1. Define the process

Name the recurring work and its boundaries.

“Improve student-success data” is broad. “Prepare the first-term progression review for one cohort” identifies work you can map and test.

**Write down:**

- What starts the process?
- What output or decision marks its end?
- Who contributes information or uses the result?
- Which part will your pilot address?

## 2. Make the aim measurable

Use the worksheet’s structure:

> By **[date]**, improve **[measure]** from **[baseline]** to **[target]**.

For example:

> Within 30 days, reduce preparation time for one course-outcome report from 90 minutes to 45 minutes, while maintaining agreement with the checked calculations.

If you do not know the baseline, measuring the current process can be your first step. Avoid choosing a target before understanding the work.

## 3. Locate the pain point

Identify the exact step, file, or handoff causing difficulty.

| Broad concern | A more specific pain point |
|---|---|
| Reporting takes too long | Someone manually retypes assessment results into a second workbook. |
| Records are inconsistent | Two files use different definitions of the same outcome. |
| Ownership is unclear | No one is responsible for updating the curriculum map after a course change. |
| The process is fragile | One faculty member knows which records to exclude, but the rule is undocumented. |
| Decisions are difficult to explain | A summary percentage omits its denominator and calculation rule. |

Specificity helps you choose an improvement that addresses the problem.

## 4. Select one principle

Choose the principle most closely connected to the pain point. Several may matter, but one can guide the first test.

| Principle | Question to ask | Possible first improvement |
|---|---|---|
| **Reproducibility** | Can someone regenerate this answer next term? | Keep the inputs and document the steps. |
| **Transparency** | Can another person understand how we produced the answer? | Define measures and make formulas and exclusions visible. |
| **Automation** | Which repeated steps do not require new judgment? | Reuse a query, linked workbook, form, template, or script. |
| **Version control** | Can we reconstruct what changed and when? | Keep versions and a change history. |
| **Data storytelling** | What does this evidence mean for an educational decision? | Explain the context, pattern, and possible action. |
| **Collaborative workflows** | Can the process continue beyond one person? | Share definitions, responsibilities, and instructions. |

## 5. Design the first small test

Describe what you will change, who will participate, and what you will retain for comparison.

For example:

> For one course report, replace manual re-entry with a reusable import process. Compare the resulting report with a checked report from the same input data.

A test can begin with a documented checklist or shared definition. It does not require a new software platform.

## 6. Choose evidence of impact

Identify **one number to track** and **one thing that must not get worse**.

| Improvement you want | Number to track | What must not get worse |
|---|---|---|
| Less repeated work | Minutes required to prepare the report | Accuracy of the calculations |
| More consistent records | Number of discrepancies between records | Time required for faculty to submit information |
| A process others can run | Steps completed independently by a colleague | Appropriate access to sensitive information |
| Faster review | Days from submission to review | Completeness of the review |
| Clearer reporting | Readers correctly interpreting a measure | Visibility of uncertainty and limitations |

Set the definitions before the pilot so that the comparison means what you intend.

# Choose an approach that fits

Coding is one way to put data science principles into practice. It can make preparation and analysis steps explicit and repeatable. The same principles can also guide work in spreadsheets, shared records, visual tools, and documented processes.

Start with the educational question and the workflow. Then choose an approach that your team can maintain.

| Approach | What it can support | Example |
|---|---|---|
| **Structured spreadsheets** | Consistent fields, visible calculations, and reusable summaries | A course-report template with validation and documented formulas |
| **Forms and shared records** | Consistent collection and an authoritative record | A form that collects required information into a shared record |
| **Saved queries and visual data preparation** | Repeatable importing, cleaning, and combining | Refreshing a saved preparation process with the next term’s data |
| **Dashboards** | Repeated access to defined measures and relevant comparisons | A report showing trends by cohort with visible denominators |
| **Workflow automation** | Routine routing, reminders, and status updates | A submission that initiates a review notification |
| **R or Python scripts** | Explicit, reusable preparation and analysis steps | A script that generates the same report from updated inputs |
| **Process documentation** | Shared definitions, roles, and checks | Instructions that let a colleague repeat a review process |

These approaches can be combined. A form might collect records, a saved query might prepare them, and a dashboard might present the results.

::: {.callout-important}
## The tool does not establish the quality of the process

A spreadsheet can contain transparent, reproducible work. A sophisticated dashboard can contain unclear definitions or unreliable inputs.

Evaluate the definitions, checks, responsibilities, and interpretation alongside the technology.
:::

## Questions to guide your choice

- Can your current tools support the small test?
- Who will maintain the approach?
- Can a colleague understand and repeat the work?
- Does the approach preserve the history you need?
- Can it provide appropriate access and protect sensitive records?
- Does it reduce work overall, including work shifted to other people?

# Check the evidence before acting

The worksheet offers three checks. Use them when designing the pilot and interpreting its results.

## What kind of question is this?

| Question type | Example | Interpretation |
|---|---|---|
| **Describing** | What percentage met the benchmark this term? | Summarizes the observed records. |
| **Predicting** | Which students may need additional support next term? | Requires evidence that the prediction performs adequately for its intended use. |
| **Explaining why** | Did a curriculum change cause an improvement? | Requires a design and analysis capable of addressing alternative explanations. |

A pattern or association can suggest a question to investigate. It does not establish a cause by itself.

## Does the measure mean what we think?

Document:

- What the measure represents.
- Its calculation and denominator.
- How missing records are handled.
- Whether definitions are consistent across terms.
- What the measure leaves out.

A precisely calculated measure can still be a poor representation of the educational outcome you care about.

## Could this be noise?

Consider the number of students and the number of observations over time.

With small cohorts, a few students can produce a substantial percentage change. Look at counts alongside percentages and examine several time points when possible.

For a small workflow pilot, a faster report on one attempt is encouraging. Repeat testing can help determine whether the improvement is sustainable.

# Establish the two guardrails

## Governance

Clarify responsibilities before using the revised process.

| Responsibility | Question |
|---|---|
| Ownership | Who is accountable for the records and definitions? |
| Entry | Who contributes information and corrects errors? |
| Editing | Who can change records, and how are changes tracked? |
| Visibility | Who needs individual records or summary information? |
| Review | Who checks the evidence and authorizes action? |
| Continuity | Who keeps the process running when someone leaves? |

## Human-centered and ethical use

Consider how the process affects the people represented in the data and the people doing the work.

- Could missing information or inconsistent definitions affect some students differently?
- Can students or faculty question and correct relevant records?
- Does the output communicate limitations?
- Does a flag initiate a conversation and contextual review?
- Does the improvement create additional burden elsewhere?

For example, a downward GPA trend can prompt a supportive conversation. Interpreting it requires context. The trend alone should not determine a progression decision.

# Put your pilot plan on one page

Copy this table into a document and complete it with the people who will participate.

| Planning item | Your response |
|---|---|
| Process and boundaries | |
| Educational or institutional challenge | |
| Aim, including date, baseline, and target | |
| Exact pain point | |
| Principle guiding the test | |
| First small change | |
| Course, cohort, or cycle included | |
| Person responsible | |
| Partners and resources needed | |
| Tools or skills needed | |
| Improvement measure | |
| What must not get worse | |
| Data definitions and quality checks | |
| Access, review, and correction responsibilities | |
| Test date and review date | |
| Decision after review | |

## A practical sequence

| When | Focus |
|---|---|
| Before the test | Confirm the aim, baseline, definitions, participants, and guardrails. |
| During the test | Record time, discrepancies, questions, and unexpected work. |
| At the review | Compare results with the aim and check what must not get worse. |
| After the review | Continue, revise, or stop the change. Document why. |

Match the timeline to the work. A process that occurs once per term may need a historical dataset or a limited trial before the next full cycle.

# Learn from the test

Use the review to answer:

1. Did the process improve in the way we intended?
2. Did accuracy, access, workload, or another protected condition worsen?
3. Could another person complete the process?
4. Which problems remained?
5. What evidence would justify expanding the change?

| Decision | When it may fit |
|---|---|
| **Continue** | The test met its aim without worsening the protected condition. |
| **Revise and retest** | The approach helped, but a definition, step, or responsibility needs adjustment. |
| **Stop or change direction** | The approach introduced unacceptable problems or did not address the pain point. |

Keep a short record of what changed, what you learned, and the next decision. That history is part of making the improvement sustainable.

# Explore how this resource was built

This companion resource was written in Quarto, which combines document text with optional executable code. Its source is maintained in a GitHub repository.

Using a document source supports review and revision. When calculations are included, rendering can regenerate their results alongside the explanation.

[View the source repository](https://github.com/madelineratoza/From-Spreadsheets-to-Systems)

<details>
<summary>Optional: see a small example of reporting with R</summary>

## A reusable calculation

The fictional records below represent how long five reports took to prepare before and during a pilot.

The example demonstrates how a calculation and its result statement can stay together. It does not establish that an improvement would occur in another program.

```{r}
#| label: optional-pilot-example

pilot <- data.frame(
  report = paste("Report", 1:5),
  before_minutes = c(90, 85, 95, 80, 100),
  pilot_minutes = c(45, 50, 40, 45, 55)
)

# Check the basic input structure.
stopifnot(
  !anyDuplicated(pilot$report),
  all(is.finite(pilot$before_minutes)),
  all(is.finite(pilot$pilot_minutes)),
  all(pilot$before_minutes > 0),
  all(pilot$pilot_minutes >= 0)
)

mean_before <- mean(pilot$before_minutes)
mean_pilot <- mean(pilot$pilot_minutes)
minutes_saved <- mean_before - mean_pilot
percent_reduction <- 100 * minutes_saved / mean_before

knitr::kable(
  pilot,
  col.names = c(
    "Fictional report",
    "Before: minutes",
    "Pilot: minutes"
  )
)
```

```{r}
#| label: optional-pilot-result
#| echo: false
#| results: asis

cat(sprintf(
  paste0(
    "In this fictional example, average preparation time changed ",
    "from **%.1f minutes** to **%.1f minutes**, ",
    "a reduction of **%.1f minutes (%.1f%%)**.\n\n"
  ),
  mean_before,
  mean_pilot,
  minutes_saved,
  percent_reduction
))
```

A team would also need to verify calculation accuracy and check whether work shifted to someone else. These preparation times alone cannot show that every part of the workflow improved.

Change the fictional times and render again to update the table and result statement.

## What makes this useful?

The inputs, calculation, and explanation are kept in the same source. Another person can inspect the rules and rerun the example.

A spreadsheet with documented formulas could support a similar small test. Code becomes useful when it fits the task, the team’s skills, and the need for repeated work.

</details>

# Further reading

The presentation draws on scholarship about learning analytics, educational data mining, and ethical use of educational data.

The entries below retain the presentation’s reading list. Bibliographic details should be verified before reuse in a publication.

1. Shorten G. Learning analytics and the future of postgraduate medical training. *Ir J Med Sci.* 2024;193(5):2607-2609.

2. Blease C, Kharko A, Bernstein M, et al. Machine learning in medical education: a survey of the experiences and opinions of medical students in Ireland. *BMJ Health Care Inform.* 2022;29(1):e100480.

3. Carney PA, Sebok-Syer SS, Pusic MV, Gillespie CC, Westervelt M, Goldhamer MEJ. Using learning analytics in clinical competency committees: Increasing the impact of competency-based medical education. *Med Educ Online.* 2023;28(1):2178913.

4. Bojic I, Mammadova M, Ang CS, et al. Empowering health care education through learning analytics: In-depth scoping review. *J Med Internet Res.* 2023;25:e41671.

5. Romero C, Ventura S. Educational data mining and learning analytics: An updated survey. arXiv. Published online February 10, 2024. doi:10.48550/arXiv.2402.07956.

6. Alfredo R, Echeverria V, Jin Y, et al. Human-centred Learning Analytics and AI in Education: A systematic literature review. arXiv. Published online December 20, 2023. doi:10.48550/arXiv.2312.12751.

7. Marcotte K, Yang P, Millis MA, et al. Ethical considerations of using learning analytics in medical education: a critical review. *Adv Health Sci Educ Theory Pract.* 2025;30(1):87-101. doi:10.1007/s10459-024-10402-7.